The Economics of Shotgun Marriage

### 2020-02-08 main/draft.tex
Fixing the August stage

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon} 
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{Economics of Shotgun Marriage}
\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu. I thank my advisors Matthias Doepke, Martí Mestieri and David Berger for outstanding guidance and proofreading as part of my dissertation committee at Northwestern. I am grateful to Alessandra Voena for fantastic amount of help and support. I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.}\\
{\small Northwestern University}}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage. I show that in American Community Survey data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future.
%This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. 
This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock,
%differences in marriage quality,
as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
\end{abstract}

@@

%[Paragraph 1: is shotgun marriage still a thing? Abstract - 2]
Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage. And, although it seems that abortion and contraception access would provide people with more choice, in this paper I show that in the data such couples still look like something pushed them to marry each other. In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces. This paper has two main goals. First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups. Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of variation in divorce rates is not just a household behavior question. For example, the divorces have large major impact on how people raise their children. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

%[Paragraph 3: general importance? Other ideas]
In addition to child development impact of unstable marriages, divorces have other important consequences. For instance, they appear as an important driver of labor market choices and outcomes of women. Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility and marriage rates. Finally, the question of marriage selection and modeling divorces is curious by its own from prospectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household questions, that is relevant for developing realistic household models and understanding things that are not easy to observe directly.

%[Paragraph 4: general importance? We need to understand trends]
On more aggregate level, this paper can broaden the understanding of recent trends in household formation. First, around 40\% of new births in 2018 occurred by unmarried mothers. Second, increase in non-marital cohabitation is a documented trend, yet the median duration of cohabitation is 22 months\footnote{Source: Copen at al,  \cite{copen}}. Despite of this, unmarried unions with children are pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers. Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births. Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly consider cohabitation that seems to become a substitute for married unions.

%[Paragraph 5: empirics --- what data do I use]
I use American Community Survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics. One important drawback of it is that the marriage history in ACS is not reported perfectly, as it only has records of the year of the most recent marriage. I overcome this by conditioning on women who was married exactly once. I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates. Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.

%[Paragraph 6: empirics --- methodological issues shortly]
My main empirical contribution is describing association between having shotgun marriages and having a divorce, that I have not found to be explicitly discussed in the existing literature. Methodologically, the crucial comparison involves partitioning population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (``marriage first'') and those who gave their first birth before or at the year of their marriage (``kids first''). I refer to the latter group as those who experienced a shotgun marriage. I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than the child's father.

%[Paragraph 7: empirica, findings --- main results]
The most important empirical result I show is substantially larger share of divorced people in ``kids first'' group. Although this difference is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$). This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school. When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.  %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.

%[Paragraph 8: empirical findings --- duration, step-children, lf and child development outcomes]
Investigating the data further leads to few additional insights. First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference. Second, the ``extra'' divorces we observe are concentrated around the time the child starts going to school. Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage. [SIPP has better data to confirm these findings?]

%[Paragraph 7: model --- setup and why do we need a model]
These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result. The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.

%[Paragraph 8: model --- what generates the result in theory]
The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality. Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally.

@@

%[Paragraph 10: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers few nontrivial counterfactual predictions exploiting interaction between unplanned pregnancies and marriage quality. First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

@@

%[Paragraph 12: what I learned about channels]
The main causal insight is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.

%[Paragraph 13: not a paragraph, just table of contents]
I elaborate on these points in the following order. Section 1 of this paper discuss existing results and general trends that are already described by the literature. Section 2 discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 4 and the process of setting its parameters in Section 5. Section 6 describes estimation results and counterfactual simulations implied by these results. Section 7 provides more general discussion and concludes.

@@

This section presents key empirical facts and several exercises on testing their validity. In summary:
\begin{enumerate}
\item Women who had their first child before their first marriage (K$\to$M) are more likely to be divorced than those who had their first child in their first marriage (M$\to$K).
\item Relative difference in the share of divorced between K$\to$M and M$\to$K is increasing with income and education
\item These results are statistically significant and robust to various methods of controlling for observable composition of the groups
\end{enumerate}

@@

General strategy involves dividing sample in two groups: couples who had kids before their marriage ($K\to M$) and those who had kids after their marriage ($M\to K$). After this division I perform multiple comparisons of these groups: the main comparison is looking at divorce rates in several variations, few additional comparisons involve marriage inequality and outcomes of children.

Although the couples are the primary focus of the paper, the surveys I use look at individuals and households. Therefore I follow females rather than males: upon divorce, children are more likely to stay with mothers (share of single mothers in ACS is ..., single fathers ...). The timing I consider in the  most general form is (for females, given both events occurred): 
\[\Delta T = T^{\text{First child  is born}} - T^{\text{First marriage happened}},\]
where $T$ refers to moments of time. After that I define two groups (given $\Delta T$ is measured in years):
\begin{align*}&M\to K \Leftrightarrow \Delta T \in \{1,2,...\},\\
&K\to M \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}

@@

All mean and quantile calculations are performed using person weights.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\linewidth]{shrs.eps}
\begin{tabular}{|p{0.65\linewidth}|}
\hline
\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$\\
\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.\\\hline
\end{tabular}
\caption{Share of K$\to$M couples and share of divorced, ACS\label{shrs}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\linewidth]{shrs-sipp.eps}
\caption{Share of K$\to$M couples and share of ever divorced, SIPP\label{shrs-sipp}}
\end{figure}

@@

This alone is not enough to draw strong conclusions: composition of these groups is drastically different. For instance, as suggested by the table, people with lower income are more likely to get divorced, so even conditional on income/education group K$\to$M couples may be poorer with each subgroup. Additionally, many other factors may drive the differences, including different ages of children (people in K$\to$M group by construction have older babies after any given number of years after marriage), regional and racial differences in attitudes to divorce, age composition, possible presence of step-children in K$\to$M couples and so on. I attempt to control for few of these things and argue that they explain these differences only partially, that suggests the presence of causal mechanism that I describe.

\subsection{Adjusting for Composition\label{adcomp}}
As people in K$\to$M and M$\to$K groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition. Appendix \ref{comp-diff-appendix} quantifies these differences, shortly, people in K$\to$M group are younger, less educated, have kids earlier, marry slightly earlier and have more unequal marriages. In this part I control for difference in observable characteristics to see what part of the difference is attributed to the fact of being in K$\to$M group. Shortly, controlling for observables mitigate some part of the difference but it still remains sizable and significant.

In general, I perform the following exercise: I regress indicator of female being divorced on set of controls $X$ and dummy variable $\I(\text{K$\to$M})_i$. The specification I use is
\[D_i = \Delta \cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i.\]
Replacing $X_i$ with a constant corresponds to raw mean comparison described above.

@@

Appendix \ref{extra-comparisons} shows few additional graphs and exercise. I summarize their results here:
\begin{enumerate}
\item In the sample I use M$\to$K women are married for longer at the moment of survey\footnote{This does not contradict the fact that K$\to$M couples marry slightly earlier as I use the sample of married women.} (so they had \emph{more} years to break up), comparing marriages of the same duration only exacerbates the difference.
\item Children of K$\to$M women are on average older (as they are born in earlier ages). Divergence between K$\to$M and M$\to$K observable divorce probabilities happens around ages 5--10. After controlling for composition, however, I show that K$\to$M couples divorce more often even conditional on child's age.
\item To further address concerns about duration, I recompute the numbers using getting the divorce at the year of survey, corresponding to marginal divorce probability as opposed to cumulative divorce probability (being divorced at by the moment survey). The magnitudes and patterns are very similar, though the numbers are harder to interpret. This suggests that 
\item I show that the difference is not likely to be driven by presence of stepchildren in K$\to$M families. In fact, as I restrict $\Delta T \geq -5$, share of stepchildren is relatively small, and surviving couples with stepchildren do not differ in observable marriage duration with couples without stepchildren.
\end{enumerate}

\subsection{Intra-Household Allocations [TBD]}
I document few additional facts describing the difference between K$\to$M and M$\to$K besides being divorced. Following Lafortune et al, 2018, I construct measure of children's grade retention based on ACS (that is an indicator if any school-age child in the household is in grade that is lower than prescribed by her/his age). I show this together with few other intra-houeshold characteristics: measure of inequality and labor force participation.

See Table \ref{intra} for the numbers. The exercise suggests that children on K$\to$M women repeat grades significantly more often, and this happens even in higher-income households. [This table is based on old sample so it is slightly inconsistent with what I do above, I will redo this in one of the future iterations.]

>>>>

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon} 
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{The Economics of Shotgun Marriage}
\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu. I thank my advisors Matthias Doepke, Martí Mestieri and David Berger for outstanding guidance and proofreading as part of my dissertation committee at Northwestern. I am grateful to Alessandra Voena for fantastic amount of help and support. I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.}\\
{\small Northwestern University}}
\date{August, 2019}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Many couples marry shortly after woman gets pregnant, which is often referred to as a shotgun marriage. I show that in American Community Survey data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future.
%This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. 
This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock,
%differences in marriage quality,
as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
\end{abstract}

@@

%[Paragraph 1: is shotgun marriage still a thing? Abstract - 2]
The term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage. And, although it seems that abortion and contraception access would provide people with more choice, in this paper I show that in the data such couples still look like something pushed them to marry each other. In particular, their marriages appear less stable over time: despite of being less than a quarter of the population shotgun marriage couples account for more than one-third of divorces.

This paper has two main goals. First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups. Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of a variation in divorce rates is not just a household behavior question. For example, divorces have major impact on how people raise their children. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

%[Paragraph 3: general importance? Other ideas]
In addition to the child development impact of unstable marriages, divorces have other important consequences. For instance, they appear as an important driver of labor market choices and outcomes of women. Additionally, being divorced with young children is considered as lowering re-marriage prospectives, and this can affect total fertility and marriage rates. Finally, the question of marriage selection and modeling divorces is interesting on its own from perspectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household decisions and to develop realistic household models.

%[Paragraph 4: general importance? We need to understand trends]
On a more aggregate level, this paper can broaden the understanding of recent trends in household formation. First, around 40\% of new births in 2018 occurred by unmarried mothers. Second, increase in non-marital cohabitation is a documented trend, yet the median duration of cohabitation is 22 months\footnote{Source: Copen at al,  \cite{copen}}. Despite of this, unmarried unions with children are pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers. Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births. Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly consider cohabitation that seems to become a substitute for married unions.

%[Paragraph 5: empirics --- what data do I use]
I use the American Community Survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics. One important drawback of it is that the marriage history in the ACS is not reported perfectly, as it only has records of the year of the most recent marriage. I overcome this by conditioning on women who was married exactly once. I then use the separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of divorce rates. Although SIPP provides superior coverage of the issues I study, the current work mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.

%[Paragraph 6: empirics --- methodological issues shortly]
My main empirical contribution is describing the association between having shotgun marriages and having a divorce. Methodologically, the crucial comparison involves partitioning population of women who were married and had kids on two groups:  those who gave their first birth the earliest during the year after their marriage (``marriage first'') and those who gave their first birth before or at the year of their marriage (``kids first''). I refer to the latter group as those who experienced a shotgun marriage. I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone other than the child's father.

%[Paragraph 7: empirica, findings --- main results]
The most important empirical result I show is substantially larger share of divorced people in the ``kids first'' group. Although this difference is not a causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$). This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school. When controlling for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares declines by about 50\%, but remains large and statistically significant.  %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.

%[Paragraph 8: empirical findings --- duration, step-children, lf and child development outcomes]
Investigating the data further leads to additional insights. First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference. Second, the ``extra'' divorces we observe are concentrated around the time the child starts going to school. Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage. %[SIPP has better data to confirm these findings?]

%[Paragraph 7: model --- setup and why do we need a model]
These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result. The model is based on a growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.

%[Paragraph 8: model --- what generates the result in theory]
The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, the couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, the woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality. Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally.

@@

%[Paragraph 10: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers predictions exploiting interaction between unplanned pregnancies and marriage quality. First, a major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, the general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

@@

%[Paragraph 12: what I learned about channels]
The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in the creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.

%[Paragraph 13: not a paragraph, just table of contents]
I elaborate on these points in the following order. Section 1 of this paper discusses the existing literature. Section 2 discusses how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain the theoretical model in Section 4 and the process of setting its parameters in Section 5. Section 6 describes estimation results and counterfactual simulations implied by these results. Section 7 provides more discussion and concludes.

@@

This section presents key empirical facts and several exercises on testing their validity. In summary:
\begin{enumerate}
\item Women who had their first child before their first marriage (kids-first) are more likely to be divorced than those who had their first child in their first marriage (marriage-first).
\item Relative difference in the share of divorced between kids-first and marriage-first is increasing with income and education
\item These results are statistically significant and robust to various methods of controlling for observable composition of the groups
\end{enumerate}

@@

General strategy involves dividing sample in two groups: couples who had kids before their marriage (kids-first) and those who had kids after their marriage (marriage-first). After this division I perform multiple comparisons of these groups: the main comparison is looking at divorce rates in several variations, few additional comparisons involve marriage inequality and outcomes of children.

Although the couples are the primary focus of the paper, the surveys I use look at individuals and households. Therefore I follow females rather than males: upon divorce, children are more likely to stay with mothers (share of single mothers in ACS is ..., single fathers ...). The timing I consider in the  most general form is (for females, given both events occurred): 
\[\Delta T = T^{\text{First child  is born}} - T^{\text{First marriage happened}},\]
where $T$ refers to moments of time. After that I define two groups (given $\Delta T$ is measured in years):
\begin{align*}&\text{Kids-first} \Leftrightarrow \Delta T \in \{1,2,...\},\\
&\text{Marriage-first} \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}

@@

All mean and quantile calculations are performed using person weights provided by the surveys.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\linewidth]{shrs.eps}
\begin{tabular}{ p{\linewidth} }
%\hline
\footnotesize \emph{Notes}:``K then M'' is kids-then-marriage and refers to women whose year of marriage is 0 to 5 years after their year of the first birth  (kids-first). ``M then K'' is marriage-then-kids and refers to women whose year of marriage is 1 to 10 years after their year of the first birth (marriage-first). Women outside of these two groups are excluded. Only women who have exactly one marriage are considered. Income refers to residualized per-hour income for subsample of working women, as described in Section \ref{inc-part}. Right panel shows share of divorced women \emph{conditional} on being in each group. ACS is American Community Survey, 2008--2016.\\
%\hline
\end{tabular}
\caption{Share of kids-first and marriage-first women and share of divorced (and not remarried) in each group, ACS data\label{shrs}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\linewidth]{shrs-sipp.eps}
\begin{tabular}{ p{\linewidth} }
%\hline
\footnotesize \emph{Notes}: Same definitions as Figure \ref{shrs} apply, except that all women who have at least one marriage are considered. SIPP is Survey of Income And Program Participation, 2014 (wave 1).\\
%\hline
\end{tabular}
\caption{Share of kids-first and marriage-first women and share of ever divorced in each group, SIPP\label{shrs-sipp}}
\end{figure}

@@

This alone is not enough to draw strong conclusions: composition of these groups is drastically different. For instance, as suggested by the table, people with lower income are more likely to get divorced, so even conditional on income/education group K$\to$M couples may be poorer with each subgroup. Additionally, many other factors may drive the differences, including different ages of children (people in kids-first group by construction have older babies after any given number of years after marriage), regional and racial differences in attitudes to divorce, age composition, possible presence of step-children in kids-first couples and so on. I attempt to control for few of these things and argue that they explain these differences only partially, that suggests the presence of causal mechanism that I describe.

\subsection{Adjusting for Composition\label{adcomp}}
As people in kids-first and marriage-first groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition. Appendix \ref{comp-diff-appendix} quantifies these differences, shortly, people in kids-first group are younger, less educated, have kids earlier, marry slightly earlier and have more unequal marriages. In this part I control for difference in observable characteristics to see what part of the difference is attributed to the fact of being in kids-first group. Shortly, controlling for observables mitigate some part of the difference but it still remains sizable and significant.

In general, I perform the following exercise: I regress indicator of female being divorced on set of controls $X$ and dummy variable $\I(\text{K$\to$M})_i$. The specification I use is
\[\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + X_i\gamma + \varepsilon_i.\]
Replacing $X_i$ with a constant corresponds to raw mean comparison described above.

@@

Appendix \ref{extra-comparisons} shows few additional graphs and exercise. I summarize their results here:
\begin{enumerate}
\item In the sample I use marriage-first women are married for longer at the moment of survey\footnote{This does not contradict the fact that kids-first couples marry slightly earlier as I use the sample of married women.} (so they had \emph{more} years to break up), comparing marriages of the same duration only exacerbates the difference.
\item Children of kids-first women are on average older (as they are born in earlier ages). Divergence between kids-first and marriage-first observable divorce probabilities happens around ages 5--10. After controlling for composition, however, I show that kids-first couples divorce more often even conditional on child's age.
\item To further address concerns about duration, I recompute the numbers using getting the divorce at the year of survey, corresponding to marginal divorce probability as opposed to cumulative divorce probability (being divorced at by the moment survey). The magnitudes and patterns are very similar, though the numbers are harder to interpret. This suggests that 
\item I show that the difference is not likely to be driven by presence of stepchildren in kids-first families. In fact, as I restrict $\Delta T \geq -5$, share of stepchildren is relatively small, and surviving couples with stepchildren do not differ in observable marriage duration with couples without stepchildren.
\end{enumerate}

\subsection{Intra-Household Allocations [TBD]}
I document few additional facts describing the difference between kids-first and marriage-first besides being divorced. Following Lafortune et al, 2018, I construct measure of children's grade retention based on ACS (that is an indicator if any school-age child in the household is in grade that is lower than prescribed by her/his age). I show this together with few other intra-houeshold characteristics: measure of inequality and labor force participation.

See Table \ref{intra} for the numbers. The exercise suggests that children of kids-first women repeat grades significantly more often, and this happens even in higher-income households. [This table is based on old sample so it is slightly inconsistent with what I do above, I will redo this in one of the future iterations.]

\begin{table}
\centering
\begin{tabular}{l  c c  c c }
\cline{2-5}
& \multicolumn{2}{ c  }{\footnotesize \textit{All couples}}  & \multicolumn{2}{ c }{\footnotesize \textit{Top 40\% of income} } \\\hline
&                       \footnotesize  Marriage-first & \footnotesize Kids-first &  \footnotesize  Marriage-first & \footnotesize  Kids-first \\\hline
\footnotesize \textit{Female in Labor Force}        & $70.3$  & $71.4$  & $79.3$ & $85.3$  \\\hline
\footnotesize \textit{Earninigs Ratio}    &  $95.2$ & $93.9$  &  $95.8$ &  $96.4$ \\\hline
\footnotesize \textit{Kids' Grade Retention}  & $\phantom{0}4.6$    & $\phantom{0}7.6$ &   $\phantom{0}3.3$   & $\phantom{0}5.9$ \\\hline\hline
\end{tabular}
\caption{Child's outcomes and intra-household allocations, all numbers are in percents.\label{intra}} %[This is borrowed from an old draft so its methodology is slightly inconsistent, to be fixed later]\label{intra}}
\end{table}
